Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-22 conclusions.tex
lloid_paper: update intro text to match figures; tightened intro and conclusion
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1134 ccdd05d5-56c1-461a-bf98-85f6876523b4

\CBC\ events may be the progenitors of some short hard gamma-ray bursts and are
expected to be accompanied prompt electromagnetic signals.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  The detection algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} data acquisition records science data in 1/16~s
blocks~\citep{Bork2001}. Data must also be aggregated from all of the \GW\
observatories using global computer networks capable of high bandwidth but perhaps only
modest latency.  This could introduce a latency of $\sim$100~ms.  Lastly, unless a
real-time infrastructure is adopted post data acquisition, it is likely that there will be 
an inherent latency introduced by the software architecture itself.  We have shown 
prototype implementation of \lloid{} using the open source signal processing platform
\gstreamer. Significant work would have to be done in order to
improve upon the latency of our implementation, including tighter integration between
data acquisition and analysis. This could be considered for third-generation detector
design.

@@

Although we have demonstrated a feasible method for advance detection we have
not yet explored triggering, coincidence, and ranking of gravitational wave
candidates in a framework that supports early \EM\ followup.  If adjacent time slices
have non-negligible cross-correlation with each other, then there may be some
subtleties in the design of these postprocessing stages that we must address in order
to translate the early-warning outputs into candidate events.  The time-slice
decomposition and the \SVD\ will help us form low-latency signal based vetoes
(for example, a $\chi^2$ statistic) that will help us here.

Our future work must also address sky localization accuracy prior to merger as well as
observing strategies.  \citet{Fairhurst2009} derives the area of 90\% localization
confidence in terms of timing uncertainties, but it would be advantageous to use
a galaxy catalog to inform the telescope tiling \citep{galaxy-catalog}.  Because
early detection is likely to be especially effective for nearby sources, the galaxy
catalog technique might be an important ingredient in reducing the fraction of
sky that must be imaged.  Extensive injection campaigns and simulations
incorporating realistic binary merger rates will be necessary in order to fully
understand the prospects for early-warning detection, localization, and \EM\ folluwup
using the techniques we have described.

\editorial{nvf: It's a bit pie-in-the-sky, but we could also mention
reconfiguring the signal recycling mirror to optimize SNR at merger. There's a
lot of science there.}

>>>>

\CBC\ events may be the progenitors of some short hard \GRB{}s and are
expected to be accompanied prompt \EM{} signals.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an \EM{} counterpart in bands from gamma-rays down to
radio.

The detection algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} data acquisition records science data in 1/16~s
blocks~\citep{Bork2001}. Data must also be aggregated from all of the \GW\
observatories using global computer networks capable of high bandwidth but perhaps only
modest latency.  This could introduce a latency of $\sim$100~ms.  Lastly, unless a
real-time infrastructure is adopted post data acquisition, it is likely that there will be 
an inherent latency introduced by the software architecture itself.  We have shown 
prototype implementation of \lloid{} using the open source signal processing platform
\gstreamer. Significant work would have to be done in order to
improve upon the latency of our implementation, including tighter integration between
data acquisition and analysis. This could be considered for third-generation detector
design.
Also possible for third-generation instruments, the \lloid{} method could
provide the input for a dynamic tuning of detector response via the signal
recycling mirror to match the frequency of maximum sensitivity to the
instantaneous frequency of the \GW{} waveform; this challenging technique has
the potential for vast gains in \SNR{} and timing uncertainty
\citep{PhysRevD.47.2184}.

@@

Although we have demonstrated a feasible method for advance detection we have
not yet explored triggering, coincidence, and ranking of gravitational wave
candidates in a framework that supports early \EM\ followup.  If adjacent time slices
have non-negligible cross-correlation with each other, then there may be some
subtleties in the design of these postprocessing stages that we must address in order
to translate the early-warning outputs into candidate events.  The time-slice
decomposition and the \SVD\ will help us form low-latency signal-based vetoes
(for example, a $\chi^2$ statistic) that will help us here.

Our future work must more deeply address sky localization accuracy prior to
merger as well as observing strategies. Here, we have followed
\citet{Fairhurst2009} in estimating the area of 90\% localization confidence in
terms of timing uncertainties alone, but it would be advantageous to use a
galaxy catalog to inform the telescope tiling \citep{galaxy-catalog}. Because
early detections will arise from nearby sources, the galaxy catalog technique
might be an important ingredient in reducing the fraction of sky that must be
imaged.  Extensive simulation campaigns incorporating realistic binary merger
rates and detector networks will be necessary in order to fully understand the
prospects for early-warning detection, localization, and \EM\ folluwup using
the techniques we have described.

### 2011-06-22 introduction.tex
lloid_paper: update intro text to match figures; tightened intro and conclusion
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1134 ccdd05d5-56c1-461a-bf98-85f6876523b4

% Outline:
%  - Low-mass inspirals are the most promising source of gravitational radiation for advanced \textsc{ligo}
% - Motivation for high throughput
%   - Point out challenges of advanced detectors: more templates, longer templates
% - Motivation for low latency
%   - Observational applications of near-real-time detection
%   - Mention S6 and VSR2, online running, EM followup
% - Layout of paper: description of method, procedure followed

The coalescence of compact binary systems consisting of neutron stars
 (\textsc{ns}) and/or black holes (\textsc{bh}) is presently the best
understood and most actively pursued potential source of gravitational
radiation for ground-based gravitational-wave (\GW{}) detectors.
As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{})
counterpart~\citep{shibata:2007}. Observation of an \EM{} counterpart to a
candidate \GW{} event will vastly boost confidence in the \GW{} detection and
will often provide sufficient sky localization to determine
the source's host galaxy and therefore measure the source's redshift. The
\GW{} observation of this event will also provide a luminosity
distance measurement. With both quantities, we can
immediately measure the Hubble constant with unprecedented
accuracy and without systematics dependent on the standard cosmic distance
ladder~\citep{2010ApJ...725..496N}. The most efficient means of obtaining
such joint observations is through the rapid identification and dissemination
of candidate \GW{} events.

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in order
to trigger followup observations by \EM{} telescopes.  The typical latencies
were 30--60 minutes ~\citep{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{} emission. We have the
ambition of reporting \GW{} candidates not minutes \emph{after} the merger, as
has already been accomplished, but seconds \emph{before}.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for several
classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s)~\citep{Lee:2005, nakar07}.
In this scenario, prompt \EM{} emission arises as shells of
relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
accompanying optical flash~\citep{Sari99}. Prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes have only been observed for a handful of long
\GRB{}s~\citep{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\citep{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly.
Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows in any band. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\citep{0004-637X-723-2-1711}.

@@

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\citep{HugheyGWPAW2011, S6lowlatency}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\citep{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

@@

\EM{} followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}. Figure~\ref{fig:sky-localization-accuracy} estimates the
minimum 90\% confidence area versus time for a few exceptionally loud
coalescence events detectable by Advanced \LIGO{} and Advanced Virgo. A
once-per-year event would have minimum 90\% confidence areas of $2250$~deg$^2$
($X$\% of the sky) and $270$~deg$^2$ ($Y$\% of the sky) at times 30~s and 10~s
before coalescence, respectively. The confidence region shrinks quadratically in
area as the overall amplitude of the signal increases. By the time of merger,
the minimum 90\% confidence area shrinks to $0.82$~deg$^2$.

>>>>

As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{})
counterpart \citep{shibata:2007}. One way of obtaining such joint observations
is through the rapid identification and dissemination of candidate \GW{} events
for \EM{} followup.

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in order
to trigger followup observations by \EM{} telescopes.  The typical latencies
were 30--60 minutes \citep{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{} emission. We have the
ambition of reporting \GW{} candidates not minutes \emph{after} the merger, as
has already been accomplished, but seconds \emph{before}.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for several
classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt \EM{} emission arises as shells of
relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
accompanying optical flash \citep{Sari99}. Prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes have only been observed for a handful of long
\GRB{}s \citep{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field \citep{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly.
Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows in any band. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
a truly exceptional event, perhaps even a glimpse of a
precursor \citep{0004-637X-723-2-1711}.

@@

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega \citep{HugheyGWPAW2011, S6lowlatency}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko \citep{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

@@

\EM{} followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}. Table~\ref{fig:earlywarning} and Fig.~\ref{fig:sky-localization-accuracy} estimate the
minimum 90\% confidence area versus time for a few rare, loud
coalescence events detectable by Advanced \LIGO{} and Advanced Virgo. A
once-per-year event would have minimum 90\% confidence areas of $1300$~deg$^2$
($3.1$\% of the sky) and $260$~deg$^2$ ($0.63$\% of the sky) at times 25~s and
10~s before coalescence, respectively. The confidence region shrinks
quadratically in area as the overall amplitude of the signal increases. By the
time of merger, the minimum 90\% confidence area shrinks to $0.82$~deg$^2$.
